\cleardoublepage

# (APPENDIX) Appendix {.unnumbered}

# Code availability

All code that we used in each phase of the project can be located on
Github. Below we listed the repositories used as well as a brief
description of their content.

```{r github-repositories-used}
tibble::tribble(
  ~Repository, ~Description,
  "<https://github.com/ccg-esb-lab/uJ>", "It contains a series of programs in $\\mu \\mathrm{J}$, which consist of an $ImageJ$ macro library for quantifying unicellular bacterial dynamics in microfluidic devices. Besides, it includes all the Python code used for the image analysis processing and our developed custom Napari cell-viewer (see Chapter \\@ref(image-processing)). ",
  "<https://github.com/jvelezmagic/undergraduate_research_project>", "It contains all the files necessary to reproduce this document in its entirety. In addition, it includes the code used in R to analyze the tabular data of the experiments (see Chapter \\@ref(experiment-analysis)).",
  "<https://github.com/jvelezmagic/CellFilamentation>", "In includes all the Julia code used to create the mathematical filamentation model exposed in Chapter \\@ref(model-analysis)."
) |>
  knitr::kable(
    format = "pipe",
    caption = "Github repositories used for this project.",
    padding = 1
  )
```

# Software tools

## Python

Below is the main list of packages used for Chapter
\@ref(image-processing)

-   Python [@10.5555/1593511].

-   dask [@rocklin2015dask].

-   ipython [@perez2007ipython].

-   matplotlib [@hunter2007matplotlib].

-   napari [@sofroniew2021a].

-   networkx [@hagberg2008exploring].

-   numpy [@2020NumPy-Array].

-   pandas [@mckinney2010data].

-   pickle [@van1995python].

-   scikit-image [@vanderwalt2014].

-   shapely [@shapely2007].

## R

Below is the main list of packages used for Chapter
\@ref(experiment-analysis) as well for the reproducibility of this
undergraduate research project.

```{r used-packages}
pkgs <- c(
  "bookdown",
  "rmarkdown",
  "knitr",
  "GGally",
  "embed",
  "fs",
  "ggdist",
  "ggpubr",
  "here",
  "janitor",
  "patchwork",
  "plotly",
  "tidymodels",
  "tidytext",
  "tidyverse",
  "stringr",
  "sessioninfo"
)
```

```{r write-packages-citations}
knitr::write_bib(
  c(
    .packages(),
    pkgs
  ),
  file = here::here("undergraduate-research-project-R-packages.bib")
)
```

```{r save-citations-vector,  results='hide'}
pkgs_citations <- knitr::write_bib(
  c(.packages(), pkgs)
)
```

```{r create-r-citations-table, results='asis'}
r_citations <- tibble::tibble(
  package = names(pkgs_citations),
  citation = pkgs_citations
) |>
  dplyr::mutate(
    citation = purrr::map(citation, as.character),
    citation = purrr::map_chr(citation, dplyr::first),
    citation = stringr::str_extract(citation, "\\{(.+)"),
    citation = stringr::str_remove_all(citation, "\\{|,"),
    citation = stringr::str_c("@", citation)
  ) |>
  dplyr::filter(stringr::str_detect(citation, "@R-")) |>
  identity()

r_citations |>
  dplyr::mutate(
    output_citation = stringr::str_c("- ", package, " [", citation, "].\n")
  ) |>
  dplyr::pull(output_citation) |>
  cat()
```

## Julia

Below is the main list of packages used for Chapter
\@ref(model-analysis).

-   Julia [@Julia-2017].
-   DrWatson.jl [@datseris2020].
-   DifferentialEquations.jl [@rackauckas2017differentialequations;
    @rackauckas2017adaptive; @rackauckas_stability-optimized_2018].
-   DataFrames.jl [@white2021].

# Software usage

## Undergraduate research project

This code base is using the `R Language` and `renv` to make a
reproducible scientific project named `undergraduate_research_project`.

1.  Clone the repository with:
    `git clone https://github.com/jvelezmagic/undergraduate_research_project`.
2.  Download latest version of [R](https://cran.r-project.org/).
3.  Open R project.
4.  Install the `renv` package with `install.packages('renv')`.
5.  Restore working environment with: `renv::restore()`.
6.  Render the book with: `bookdown::render_book()`.
7.  Edit documents and render again.

## Cell-viewer

This code base is using the `Python Language`.

1.  Clone the repository with:
    `git clone https://github.com/ccg-esb-lab/uJ`.
2.  Go to `single-channel` directory.
3.  Inside of `MGGT-AMP-Pulse` (*i.e.*, chromosome strain) or
    `pBGT-AMP-Pulse` (*i.e.*, plasmid strain) enter to
    `6_Lineages_corrector_napari.ipynb`.
4.  Change the parameters and use it.

## Filamentation model

This code base is using the `Julia Language` and `DrWatson` to make a
reproducible scientific project named `CellFilamentation`.

1.  Clone the repository with:
    `git clone https://github.com/jvelezmagic/CellFilamentation`.
2.  Download latest version of
    [Julia](https://julialang.org/downloads/).
3.  Open Julia project.
4.  Open Julia console and do the following to restore working
    environment:

``` {.julia}
using Pkg
Pkg.activate(".") # Path to the project.
Pkg.instantiate()
```

5.  Play with the model.

# Colophon

This undergraduate research project was written in
[RStudio](https://www.rstudio.com/products/rstudio/)
using [**bookdown**](https://bookdown.org).
The [website](https://jvelezmagic.github.io/undergraduate_research_project/) is
hosted via Github Pages, and the complete source is
available via Github.

This version of the project was built with `r R.version.string` and the
following packages:

```{r colophon, results='asis'}
pkgs |>
  sessioninfo::package_info(dependencies = FALSE) |>
  tibble::as_tibble() |>
  dplyr::transmute(
    Package = package,
    Version = ondiskversion,
    Source = stringr::str_replace(source, "@", "\\\\@"),
    Source = stringr::str_remove(Source, "\\\\@[a-f0-9]*")
  ) |>
  knitr::kable(
    format = "pipe",
    caption = "Packages used to built the project documents."
  )
```
